Just Take the Bus: Analysis of Toronto Bike Theft from 2022-2024

### 2024-11-10 Paper/Article.Rmd
Create Article.Rmd

---
title: "Just Take the Bus: Analysis of Toronto Bike Theft from 2022-2024"
subtitle: idk 
author:
- Rayan Awad Alim
- Samreen Gill
date: "November 4th 2024"
output: pdf_document
date-format: long
abstract: "This analysis examines Bike theft data obtained from the Toronto Police Data Portal."

format: pdf
thanks: 'Code and data are available at: https://github.com/RayanAlim/US_Census_Education_Data_Analysis/'
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(dplyr)
library(here)
library(knitr)
```

```{r}
bike_thefts_csv <- read_csv(here("Data", "Bike_thefts_raw.csv"))
head(bike_thefts_csv)
```

```{r}
library(lubridate)
bike_thefts_csv$OCC_DATE <- ymd_hms(bike_thefts_csv$OCC_DATE)
bike_thefts_csv$REPORT_DATE <- ymd_hms(bike_thefts_csv$REPORT_DATE)
```

```{r}
library(ggplot2)
library(dplyr)

bike_thefts_csv %>%
  mutate(Month = floor_date(OCC_DATE, "month")) %>%
  group_by(Month) %>%
  summarize(Count = n()) %>%
  ggplot(aes(x = Month, y = Count)) +
  geom_line(color = "blue") +
  labs(title = "Monthly Bike Theft Trends in Toronto",
       x = "Month", y = "Number of Thefts") +
  theme_minimal()

```

```{r}
ggplot(bike_thefts_csv, aes(x = BIKE_COST)) +
  geom_histogram(binwidth = 100, fill = "skyblue", color = "black") +
  labs(title = "Distribution of Stolen Bike Costs",
       x = "Cost of Bike (CAD)", y = "Frequency") +
  theme_minimal()

```

```{r}
library(leaflet)

leaflet(data = bike_thefts_csv) %>%
  addTiles() %>%
  addCircleMarkers(
    ~LONG_WGS84, ~LAT_WGS84,
    color = "red", radius = 3,
    popup = ~paste("Cost:", BIKE_COST, "<br>",
                   "Date:", OCC_DATE)
  ) %>%
  addLegend("topright", colors = "red", labels = "Bike Theft", title = "Theft Locations")

```

##

```{r}
bike_thefts_csv %>%
  count(NEIGHBOURHOOD_158) %>%
  ggplot(aes(x = reorder(NEIGHBOURHOOD_158, n), y = n)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Bike Thefts by Neighborhood",
       x = "Neighborhood", y = "Number of Thefts") +
  theme_minimal()
# NOTE TO ME: cut to TOP 10
```

```{r}
bike_thefts_csv %>%
  count(OCC_DOW) %>%
  ggplot(aes(x = OCC_DOW, y = n, fill = OCC_DOW)) +
  geom_bar(stat = "identity") +
  labs(title = "Bike Thefts by Day of the Week",
       x = "Day of the Week", y = "Number of Thefts") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set3")
```

@@

# Data

We used the statistical programming language R [@citeR], dpylr [@dpylr], here [@here], and tidyverse [@rohan] to assist with cleaning and analyzing the data. We rendered our table using knitr [@citeknitr].

The data we used is taken from IPUMS [@ipumsusa].

Possible viz:
- Neighbourhoods 
- Months of the year 
- Day of the week 
- Value 
-

Possible Extensions:
- vs car thefts

>>>>

---
title: "Just Take the Bus: Analysis of Toronto Bike Theft from 2022-2024"
subtitle: tbd
author:
- Rayan Awad Alim
- Samreen Gill
thanks: https://github.com/RayanAlim/Toronto_Bike_Thefts_Data_Analysis
date: "November 12, 2024"
abstract: This analysis examines Bike theft data obtained from the Tronrot Police Data Portal.
output: pdf_document
toc: true
number-sections: true
editor_options:
  markdown:
    wrap: 72
---

```{r setup}
#| include: false
#| warning: false
#| message: false
# Workspace Set up

knitr::opts_chunk$set(echo=FALSE, include=TRUE, eval=TRUE)
required_packages <- c(
  "lubridate", "sf", "tidyverse", "ggplot2", "maps", "dplyr", "sf", "mapdata", "ggspatial", "opendatatoronto", "here"
) 
for (p in required_packages) { # install if missing
  if (!require(p, character.only = TRUE)) {
    install.packages(p, character.only = TRUE)
  }
}

library(ggplot2)
library(maps)
library(dplyr)
library(tidyverse)
library(sf)
library(mapdata)
library(ggspatial)
library(opendatatoronto)
library(here)
```

```{r}
#| include: false
#| warning: false
#| message: false
# you are welcome to use either the CSV file or the GeoJSON file
# GeoJSON format maybe useful if you would like to combine
# an external geospatial dataset

bike_thefts <- read_csv(here("Data", "Bicycle_Thefts_Open_Data.csv")) |>
  mutate(
    occurence_date = as_date(
      OCC_DATE, format = "%a, %d %b %Y %H:%M:%S GMT"
    ),
    report_date = as_date(
      REPORT_DATE, format = "%a, %d %b %Y %H:%M:%S GMT"
    )
  )
bike_thefts_sf <- read_sf(here("Data", "Bicycle_Thefts_Open_Data.geojson")) |>
  mutate(
    occurence_date = as_date(
      OCC_DATE, format = "%a, %d %b %Y %H:%M:%S GMT"
    ),
    report_date = as_date(
      REPORT_DATE, format = "%a, %d %b %Y %H:%M:%S GMT"
    )
  )
```

```{r}
#| include: false
#| warning: false
#| message: false
#glimpse(bike_thefts)
```

## TIME TRENDS ANALYSIS

```{r}
#BY MONTH
bike_thefts %>%
  count(OCC_MONTH) %>%
  ggplot(aes(x = reorder(OCC_MONTH, n), y = n, fill = OCC_MONTH)) +
  geom_bar(stat = "identity") +
  labs(title = "Bike Thefts by Month of the Year",
       x = "Month", y = "Number of Thefts") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set3") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```

```{r}
# BY DAY OF WEEK
bike_thefts %>%
  count(OCC_DOW) %>%
  ggplot(aes(x = OCC_DOW, y = n, fill = OCC_DOW)) +
  geom_bar(stat = "identity") +
  labs(title = "Bike Thefts by Day of the Week",
       x = "Day of the Week", y = "Number of Thefts") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set3")
```

```{r}
# heaatmap of thefts by day of week and hour

ggplot(bike_thefts, aes(x = OCC_DOW, y = OCC_HOUR)) +
  geom_bin2d(bins = 24) +
  scale_fill_gradient(low = "lightblue", high = "red") +
  labs(title = "Heatmap of Bike Thefts by Day and Hour", 
       x = "Day of Week", 
       y = "Hour of Day") +
  theme_minimal()
```

### Neighbour hood analysis and Maps

```{r}
#TOP TEN 
bike_thefts %>%
  count(NEIGHBOURHOOD_158) %>%
  slice_max(n, n = 10) %>%
  ggplot(aes(x = reorder(NEIGHBOURHOOD_158, n), y = n)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 10 Neighborhoods for Bike Thefts",
       x = "Neighborhood", y = "Number of Thefts") +
  theme_minimal()
```

```{r}
#| include: false
#| warning: false
#| message: false
#MAP TBD
#library(opendatatoronto)
```

### Bike Theft by Type and Cost
Here we com[are] the frequency of thefts by bike type and cost distribution:

```{r}
# Barplot for bike types
ggplot(bike_thefts, aes(x = BIKE_TYPE)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Number of Bike Thefts by Bike Type", x = "Bike Type", y = "Count") +
  theme_minimal()
```

```{r}
# Box plot for bike cost distribution by status
ggplot(bike_thefts %>% filter(!is.na(BIKE_COST)), aes(x = STATUS, y = BIKE_COST)) +
  geom_boxplot() +
  labs(title = "Bike Cost Distribution (Recovered vs Stolen)", x = "Status", y = "Cost ($)") +
  theme_minimal()
```
### Reporting Delays
Analyzing the time delay between theft occurrence and reporting date (Not working, got to tweajk it some more)

```{r}
#| warning: false
#| message: false
# calcuate reporting delay
bike_thefts <- bike_thefts %>%
  mutate(reporting_delay = as.numeric(difftime(ymd_hms(REPORT_DATE), ymd_hms(OCC_DATE), units = "days")))

ggplot(bike_thefts, aes(x = reporting_delay)) +
  geom_histogram(binwidth = 1, fill = "purple", color = "black") +
  labs(title = "Reporting Delays for Bike Thefts", x = "Days Delay", y = "Count") +
  theme_minimal()
```

### Theft Recovery Analysis by Neighborhood
which neighborhoods have higher recovery rates and analyzing the recovered vs unrecovered bikes by neighborhood

```{r}
# calculating recovery rate by neighborhood
recovery_by_neighborhood <- bike_thefts %>%
  group_by(NEIGHBOURHOOD_158) %>%
  summarize(
    total_thefts = n(),
    recovered = sum(STATUS == "RECOVERED"),
    recovery_rate = recovered / total_thefts
  ) %>%
  slice_max(recovery_rate, n = 10) # Get top 10 neighborhoods by recovery rate

#  recovery rate by neighborhood
ggplot(recovery_by_neighborhood, aes(x = reorder(NEIGHBOURHOOD_158, recovery_rate), y = recovery_rate)) +
  geom_bar(stat = "identity", fill = "forestgreen") +
  coord_flip() +
  labs(title = "Top 10 Neighborhoods by Recovery Rate", x = "Neighborhood", y = "Recovery Rate") +
  theme_minimal()
```

@@
